\OurTool{}: Integrative Probing Analysis of Nucleic Acids Empowered by Multiple Accessibility Profiles

### 2019-12-22 Response-NAR-Round1.tex

\documentclass[9pt,hyperref]{article} % use larger type; default would be 10pt
\setcounter{secnumdepth}{1}

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

%%% Examples of Article customizations
% These packages are optional, depending whether you want the features they provide.
% See the LaTeX Companion or other references for full information.

%%% PAGE DIMENSIONS
\usepackage[margin=1.3in]{geometry} % to change the page dimensions
\geometry{a4paper} % or letterpaper (US) or a5paper or....
% \geometry{margin=2in} % for example, change the margins to 2 inches all round
% \geometry{landscape} % set up the page for landscape
%   read geometry.pdf for detailed page layout information

@@

\usepackage{graphicx} % support the \includegraphics command and options

% \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{hyperref}
\hypersetup{
	colorlinks, linkcolor={red!80!black},
	citecolor={blue}, urlcolor={red!50!black}
}

\usepackage[authoryear]{natbib}
\usepackage{amsmath} % for much better looking tables
\usepackage{xspace}
\usepackage{url}
\usepackage{soul}
\usepackage[table,x11names,dvipsnames]{xcolor}
\usepackage{amssymb} % for much better looking tables
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{ntheorem} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
\usepackage{mdframed}
\usepackage{relsize}
\usepackage{pgf}
\usepackage{cabin}
\usepackage{enumitem}
\usepackage{tgtermes}
\usepackage{collcell}
\usepackage{adjustbox}% to resize figures

%\usepackage{tgadventor}
% These packages are all incorporated in the memoir class to one degree or another...

@@

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customise the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

%%% SECTION TITLE APPEARANCE
\usepackage{sectsty}
\allsectionsfont{\sffamily\mdseries\upshape} % (See the fntguide.pdf for font help)
% (This matches ConTeXt defaults)

%%% ToC (table of contents) APPEARANCE
\usepackage[nottoc,notlof,notlot]{tocbibind} % Put the bibliography in the ToC
\usepackage[titles,subfigure]{tocloft} % Alter the style of the Table of Contents
\renewcommand{\cftsecfont}{\rmfamily\mdseries\upshape}
\renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!

\newcommand{\FAnswer}[1]{{\color{black!95}#1}}
\newcommand{\IAnswer}[1]{\FAnswer{\hfill $\to$ #1}}
\newcommand{\PAnswer}[1]{\FAnswer{\\#1}}

\newcommand{\Answer}[1]{\noindent{{{\sffamily Response:} }{\FAnswer{#1}}}}
\newcommand{\Comment}[1]{\item{\color{red!40!black}{{\sffamily Comment:} }{#1}}}
\newcommand{\Quote}[1]{\begin{mdframed}[backgroundcolor=black!2,linecolor=gray!50, linewidth=.6mm]#1\end{mdframed}}

\usepackage[colorinlistoftodos]{todonotes}
\newcommand{\TODO}[2][All]{{\todo[color=blue!10,inline]{\color{black}{TODO (#1)}: #2}}}
\newcommand{\TODOBruno}[2][Bruno]{{\todo[color=red!10,inline]{\color{black}{TODO (#1)}: #2}}}
\newcommand{\TODOAfaf}[2][Afaf]{{\todo[color=Aquamarine3,inline]{\color{black}{TODO (#1)}: #2}}}

@@

\tableofcontents
\listoftodos

\section{General comments}

First of all, the authors would like to express their deepest gratitude for the quality and thoroughness of the reports written by the reviewers, covering a remarkably comprehensive range of aspects of the submitted manuscript. We have individually addressed each of the \ref{comment:last} comments to the best of our ability, but would like to start this note with higher-level answers to recurrent questions:

\TODO[Yann]{Faire résumé}
Points géneraux :
\begin{itemize}
\item On a probablement pas été super clairs, mais :
Pas de comparaison entre protocoles, plutôt un test, rendu possible par IPANEMAP de la complémentarité des protocoles
\item Pas optimisation du traitement des donnés de sondage, mais traitement uniforme
\item Approche multi probing nous parait susceptible d’être robuste à l’incorporation d’une réactivité moins informative.
\end{itemize}

@@

\section{Detailed responses}

\subsection{Reviewer \#1}

\begin{enumerate}
	\Comment{The authors do not provide any sense of how raw data sets correlate in their experiments. They only analyze the ensembles of structures predicted based on their data. Thus for example, in Figure 5 we notice that the method of detection (CE) seems to be a far greater differentiator in predictions than the probe used (e.g. not all 1M7 cluster). The authors should add a panel clustering the raw probing data to compare it with the IPANEMAP clustering. This will require some clever design of distance metrics and mapping 1M7 to DMS may be complex, but they should do a heat map like Figure 6 but with the raw data minimally.}

	\Answer{
		First of all, we need to stress that it is not the objective of our manuscript to establish the predictive potential of various probing technologies and reagents. Rather, we wish to embrace the diversity of available protocols, and harness the resulting data to improve its integration and exploitation within structure prediction pipelines.

		That being said, we agree that a cross-comparison of the raw reactivity profiles is potentially informative, and useful to better understand the data. Accordingly, we now dedicate a whole section of the supplementary data to a correlation analysis, whose results and conclusions are reproduced below.

		\Quote{
			\subsection*{Cross-analysis of raw reactivities profiles}

			We compared raw reactivity data produced across various conditions, in order to test their overall compatibility, investigate their main determinants and compare their impacts on predictions.

						\includegraphics[width=.54\textwidth]{graphs/didy/reactivity_correlation.pdf}\hfill
			\includegraphics[width=.45\textwidth]{graphs/didy/bi_clustering.pdf}\\
			\noindent Figure~A: Correlation of reactivity profiles (left) and induced (pseudo-)ensemble distance (right). Conditions clustered in our analysis due to their impact of the pseudo-Boltzmann ensemble (black boxes, right) do not necessarily correlate very well in term of raw reactivity (white boxes, left).

			\paragraph{Correlation metrics and rationale.}To that purpose, we consider the Pearson product-moment \Def{correlation coefficients} $C_{R,R'}$ of two reactivity profile $R$ and $R'$, defined as
			$$ C_{R,R'} = \frac{\sum^n_{i=1}\left(R_i - \mu_R\right)\left(R'_i - \mu_{R'}\right)}{\sqrt{\sum ^n _{i=1}\left(R_i - \mu_{R}\right)^2} \sqrt{\sum ^n _{i=1}\left(R'_i - \mu_{R'}\right)^2}},$$ 
			where $\mu_{R}$ and $\mu_{R'}$  denote the mean reactivity within $R$ and $R'$ respectively.

			The choice of the \Def{correlation coefficient as a measurement of pairwise similarity} for probing profiles is motivated by differences in the distributions of reactivities across experimental conditions and protocols. 
			Indeed, $C_{R,R'}$ remains unchanged by the application of any rescaling, through an affine transformation with positive slope, to either $R$, $R'$ (or both). Moreover, this metrics does not scale with the number of positions.
			Thus, we can restrict the sums and mean values to positions for which reactivities are available in both experiments, and still compare the coefficient with another pair of reactivity profiles. This property is important in order to tolerate \Def{missing values} introduced by experimental artifacts and reagent properties, while still being able to compare the coefficients.

			\paragraph{Results.} We computed the correlation coefficient for all pairs of reactivity profiles, and report the results in the left matrix of Figure~A. For the sake of comparison, we also remind the Ensemble Distance matrix, indicating the difference in impact on the computational predictions of each pair of conditions.

			We first observe that almost all of the pairs of reactivity profiles show a positive correlation, as expected from their observation of a common structure. A notable exception is \OneMSevMgCE, having average negative correlation with all its peers (except for \NMIAMgCE). Its outlier status is confirmed by its large ensemble distance to other conditions, yet does not prevent it to achieve a respectable MCC of 70\%. Finally, the perfectly negative correlation between \CMCTMg{} and \DMSMg{} can be dismissed as a computational artifact, since CMCT and DMS react with different nucleotides, and the correlation cannot be computed.

			Moreover, we observe that conditions clustered from the Ensemble Distance metrics (due to their impact of the pseudo-Boltzmann ensemble) do not always correlate in term of raw reactivity, with very poor correlations occurring within clustered conditions. This essentially reflects the fact that similar profiles may induce very similar pseudo-Boltzmann ensembles, \emph{ie} different premises sometimes imply the same conclusions.

			Conversely, one observes very few strong correlations outside of clusters. In fact, the only two exceptions are \OneMSevILU and \OneMSevILUThree which, despite belonging to different clusters and achieving different qualities of prediction, correlate almost perfectly with each other and with other reactivity profiles produced using 1M7/SHAPE-MaP. Other natural candidates for explaining the reactivity correlations (Map vs CE, +Mg vs -Mg) do not appear to be supported by this analysis.

			Overall, this analysis confirms the consistency of the produced probing profiles (positive correlation), but also suggests a very delicate effect of reactivity data on the downstream predictions, with very different profiles leading to similar predictions (\OneMSevILU), and similar profiles inducing very different ones (\OneMSevILU vs \OneMSevILUThree).}

				In particular, while reactivity profiles correlate positively in a pairwise fashion, as can be expected, these results do not support the hypothesis that the method of detection (CE vs Map) would represent a strong differentiator at the level of reactivity profiles. In fact, the only clear clusters are:
		\begin{itemize}
			\item  $\{\NMIAMg,\NMIA\}$;
			\item  $\{\NMIACE,\NAICE,\OneMSevCE\}$;
			\item $\{\OneMSevILU, \OneMSevILUThree,\OneMSevILUThreeMg,\OneMSevILUMg\}$,
		\end{itemize} having respectively a similar reagent, technology, or both, as a common trait.

@@

	}

	\Comment{I am also concerned by the fact that they only have one (-) No Prob condition. For both CE and MaP data, a background subtraction is required. It is not at all clear what the No Prob condition is and why it was not independently subtracted from all the data prior to analysis.}

	\Answer{There is a misunderstanding here, the “(-) No Prob” conditions refers to a baseline computational prediction that does not use any probing data as constraints (Turner energy model only). We understand that this may be confusing to the reader, and we now refer to it as “(-) No Exp” for modeling without experimental data.

	 Of course we have carried out control experiments without any modification reageant for each of the reported experiments, as now explicitly specified in the material and methods section.
	 \TODOBruno{J'écris ça, mais je suis plus trop sûr, tu peux vérifier: \\ \didy{} was also probed with DMS (DiMethylSulfate) and CMCT, this time only in presence of Mg$^\text{2+}$, resulting in a total of 16 \Def{probing conditions}, a term we use in the following to denote a combination of probing technology, reagent and presence/absence of Mg$^\text{2+}$ (+ sequencing technology). For each probing condition, three experiments were performed in presence/absence of the reagent and in a denatured setting, following classic protocols~\cite{Wilkinson2006,Smola2015}.}}

@@

	\Comment{I am concerned the MaP data they collected is not reporting on structure. For example, quality controls need to be established showing that DMS reactivities follow expected profiles, i.e. A and C having higher mean reactivity that G and U.}

	\Answer{We understand this legitimate concern, expressed by multiple referees. The probing data reported on the structure for each probing conditions and reagents is now provided in Supplementary Figures S3 and S4. Also, please note that the DMS results were not obtained following a SHAPE-MaP-like strategy but by a “classical” RT-stop technique and Capilary electrophoresis.}

@@

	\Comment{The main reason I am concerned with their MaP data is that they used superscript III instead of II for the RT. In the Smola paper, superscript III was attempted, but it was shown it does not detect adduct formation by mutation. If the CE to MaP correlations are low for the data, which I suspect they are, the authors will have to repeat MaP profiling with Superscript II to show this change in the protocol does not affect the data. I suspect however it will and may change the conclusions of the paper.}

	\Answer{It is exact that the Weeks laboratory uses an RNAse H minus mutant of the MMLV RT sold as “Superscript II” and advises to use it. However as far as we can tell from reviewing Dr Weeks bibliography, they never reported the test of “superscript III” (another RNAse H- mutant of the MMLV RT sold by the same company), and therefore never showed that this technology does not detect the adduct.

		More precisely, in \cite{Smola2015}, the authors state that “Reaction conditions have been optimized for SuperScript II reverse transcriptase only. Other reverse transcriptase enzymes and derivatives have not been tested and should not be used”. In this work, we tried superscript III in the  described conditions, and inspected the coverage/mutation distribution graphs produced by ShapeMapper. Those graphs are suggested by the Weeks group to assess the success of a SHAPE-MaP experiment, as described in \cite{Smola2015}.

	Please find below some examples of the histograms obtained using Superscript III:

	\begin{itemize}
	\item \NMIA:

	{\centering\includegraphics[width=.8\linewidth]{graphs/didy/NMIA}\\}

	\item \NMIAMg:

	{\centering\includegraphics[width=.8\linewidth]{graphs/didy/NMIA-Mg}\\}

	\item \OneMSevILUThree:

	{\centering\includegraphics[width=.8\linewidth]{graphs/didy/1M7}\\}

	\end{itemize}

	As can be seen on the right  histograms the read depth is sufficient ($>$10k reads/nucleotides) for these experiments to be interpreted (\emph{ie} for the frequency of mutations to be a good estimate of the probability). The histograms on the left shows that there are significantly more mutations in the modified (denatured and SHAPE) samples than in the untreated ones.

	These histograms are very similar to those deemed as exemplary of a successful experiment by \cite{Smola2015},. 
	In addition, those SHAPE-MaP results are consistent with the X-ray structure (see Supplementary Figure S2 and S3), and their use with our prediction software was shown to improve the accuracy of the obtained model(s).

	In conclusion, a thorough re-inspection of the data does not lead us to question the relevance of these results, and we do not see the necessity to repeat them using superscript II.}

@@

	\Comment{One possibility if the authors indeed find the MaP data is not correct because of superscript III would be to not include it in the analysis. Unfortunately this may somewhat temper the general interest of the paper, as a majority of SHAPE and probing data is now collected with MaP and -stop protocols.}

	\Answer{As discussed above, we did not find any reason to doubt the quality of our results, and believe they should remain included in our analysis.}

@@

	\Comment{The authors also need to make their raw data available, I would recommend using the SNRNASM standard (PMID 21610212) and adding the files in the supplement. In addition the raw read files and alignments should really be deposited into the sequence read archive (SRA) so others can assess the quality of the mutation rates.}

	\Answer{We agree that raw data should be made available to the reader for the sake of reproducibility. Unfortunately, the SNRNASM resource was no longer available at the time of this revision. We have informed the creators of the resource, but have not witnessed accessibility improvement over the course of our revisions. 
	In the meanwhile, we have deposited the reactivity files on the GitHub page of IPANEMAP, where they also serve as illustrative examples.

	We experienced other issues while attempting to upload the raw data onto the SRA. Namely, this database requires a structured definition of the experiment, including several mandatory fields (growth culture...) that we could not reasonably fill given the nature of our experiments. In fact, a cursory search of the database showed that primary data produced by the Weeks lab is very rarely deposited in the SRA, indicating that similar difficulties are also encountered by other groups.

	We have permanently uploaded the data at:

	{\centering \url{http://www.lix.polytechnique.fr/~ponty/suppdata/ipanemap/}\\}

	and commit to its continued accessibility (but will also pursue public venues for better guarantees of a continued availability). 
	\TODO{Add minimal page and reactivities, including scripts/instructions to compute reactivities.}

@@

	\begin{enumerate}[resume]

	\Comment{The authors picked with DiLCrz a hard example since most problems of prediction accuracy are caused by the confusion of the nested structure model in the presence of the reactivities of the pseudoknot. That is, the quite stable PK-subhelix of length 4 (ignored in the nested model and the evaluation) will have a strong impact on reactivities that can not (correctly) be captured. Thus, the authors mainly evaluate how much the prediction of local structures in subsequence 65-120 is correct or not (as also visible in the reported structures within the supplement). Eventually, this also impacts the generality of the results.

	Thus, I think the authors should add a respective discussion to the manuscript.}

	\Answer{It is true that at least four helices of DiLCrz (DP2, DP2.1, P10 and P9) are somewhat easy to predict, and are indeed consistently found in all predictions, including in the probing-free \Software{RNAfold} MFE prediction. The changes in performances are thus mostly due to the other 6 helices (omitting the pseudoknot). We emphasized this fact in the description of our result section, to provide the reader additional context for the interpretation of our results. However, we disagree that this impacts the generality of our results. In particular, badly produced/poorly combined reactivities may easily disrupt the four above helices, and the fact that our method is able to preserve those helices, yet predict other structure elements, should be considered an asset of our method.

	The remark regarding the impact of the pseudoknot subhelix (P7) on the reactivity is very perceptive, and certainly noteworthy. Indeed, positions 111 to 115 (the 5' region of P7) are typically associated with lower reactivities, providing incentives to algorithms for pairing them within their proposed models. Accordingly, one finds those positions to be paired to erroneous partners within most of our models. Such is not the case, however, for the 3' region of P7, which is typically left unpaired in our models.
}

	\Comment{The methods section needs some reordering to first formaly introduce clusters and respective measures (stability/support) (e.g. in line 3-1-57) before introducing and discussing MBkM. For this restructuring, you might want to introduce the cummulative sample multi-set "$\mathcal{S} = \sum_d \mathcal{S}_d$" from which the clusters (also multi-sets?!) form a partitioning.}

	\Answer{We fully agree that reproducibility is essential, and have realized that our previous manuscript was lacking in details. Accordingly, we have largely revised our description of the method to clarify the nature of the clusters (indeed multisets), and the exact computation of centroids (including base-pair/unpaired position probabilities within a cluster).

### 2019-12-22 IPANEMAP-2019.tex

\maketitle
\begin{abstract}
The manual production of reliable RNA structure models from chemical probing experiments, benefits from the integration of information derived from multiple protocols and reagents. However, the interpretation of multiple probing profiles remains a complex task, hindering the quality and reproducibility of modeling efforts.

@@

SHAPE reagents represent a popular class of small molecules that react with the 2’ hydroxyl of flexible ribose~\cite{McGinnis2012}, although the exact properties observed by SHAPE remain the object of ongoing investigations~\cite{McGinnis2012,Hurst2018,Mlynsky2018,Frezza2019,Busan2019}.
As ribose flexibility is proportional to the degree of freedom of the nucleotide, it is assumed that SHAPE reagents discriminate nucleotides in stable interactions from others. Different SHAPE reagents are endowed with different dynamics, that can help differentiate Watson-Crick base pairs from more dynamics tertiary contacts~\cite{Gherghe2008,Steen2012,Rice2014,Busan2019}. Other chemical probes, harbouring different chemical reactivities, have been developed. Amongst the most popular, DiMethyl Sulfate (DMS) is a small molecule that methylates N1 of adenines and N3 of cytosine if they are not involved in a hydrogen bond. Similarly, 1-cyclohexyl-3-(2-morpholinoethyl)carbodiimide metho-p-toluene sulfonate (CMCT) reacts on the Watson-Crick face of guanosine (N1) and uracyl (N3) when not engaged in a hydrogen bond~\cite{Ehresmann1987,Brunel2000}. These two reagents therefore mostly reveal Watson-Crick base pairing, but also any other type of contact involving the same edge. This diversity of probes not only increases  coverage of the different positions and structural contexts in an RNA, but it also provides qualitatively different qualitative information, as been widely used since the very early days of RNA structure probing~\cite{Moazed1986, Ehresmann1987, Romaniuk1988, Butcher1994, Brunel2000,  Cordero2015, Somarowthu2015, Gross2017}.

@@

Computationally, the reactivity of a nucleotide is typically used as a proxy for the unpairedness of individual nucleotides. The last couple of decades have nevertheless seen a series of paradigm shifts in the ways probing information is integrated, somehow mirroring the evolution of {\em ab initio} methods for secondary structure prediction. The seminal work of Mathews~\cite{Mathews2004} used cutoffs to transform reactivity values into hard constraints. Depending on the reagent/enzyme, significantly unreactive or reactive positions were forced to either paired or remain unpaired respectively within predicted structures. However, such \emph{hard} constraints turned out to be very sensitive to the choice of the cutoff, easily leading to artificially unpaired predictions or unsatisfiable sets of constraints. With the advent of SHAPE probing, subsequent methods lifted the requirement of thresholds and integrated reactivities directly as pseudo-energies, supplementing the Turner nearest-neighbor energy model~\cite{Turner2010} within an (pseudo-) energy minimization scheme~\cite{Deigan2009,Zarringhalam2012,Washietl2012}. Such methods can be seen as optimizing a tradeoff between the thermodynamic stability of a model, assessed by the Turner energy model, and its compatibility with the reactivity profile. As a consequence, popular packages for secondary structure prediction, such as \Software{RNAStructure}~\cite{Mathews2004} or the \Software{Vienna} package~\cite{Lorenz2011}, now routinely accept SHAPE-like reactivity profiles as input of their main structure prediction methods. These include the joint folding of aligned RNAs~\cite{Lavender2015}, partition function, statistical sampling and Maximal Expected Accuracy predictions~\cite{Spasic2017}\ldots

However, independent computational predictions for the same RNA using probing data obtained with different probes often yield largely different models, none of which are fully consistent with all the probing data. It follows that, while theoretically informative, a multiprobing strategy often leaves a user with different models, from which one cannot objectively decide. In order to identify the best fit with all the data, researchers resort to very intuitive and manual methods such as projecting the results obtained with one probe onto the different models obtained using the constraints from another probe~\cite{Herbreteau2005,James2008,Weill2004}. At the end of the modeling process, the modeler is often left with several alternatives, all of which may appear equally consistent with the probing data, and the choice of a "best" model becomes somewhat subjective. Here we have developed a new modeling procedure that jointly takes into account multiple probing data, and ultimately yields a small collection of secondary structure models.

@@

\subsection*{The \OurTool{} method}
We introduce the Integrative Probing Analysis of Nucleic Acids Empowered by Multiple Accessibility Profiles (\OurTool{}), a novel approach that integrates the signal produced by various chemical probes.

@@

	\caption{\OurTool{} workflow: \OurTool{} takes as input an RNA sequence with profiling data, denoted by reactivities, from various experimental conditions. \OurTool{} proceeds, first,  with a stochastic sampling that results into samples of predicted secondary structure. The data-driven predicted structures are then gathered in one sample, serving as input for the clustering step. \OurTool{} proceeds, then, with an iterative clustering that ends once the stopping criterion are reached. This step allows to identify the adequate number of clusters $k$ to be considered. The $k$ resulting clusters are then identified by their centroid structures. Clusters figuring on the 2D-Pareto frontier are considered to be optimal and subsequently their corresponding centroids are reported as the predicted structure through \OurTool.}\label{fig:approach}
\end{figure}

Our method, summarized in Figure~\ref{fig:approach}, takes as input a set $\mathcal{D}$ of probing experiments, each materialized by a reactivity profile. It starts by producing sets of representatives structures for each of the reactivity profiles using a SHAPE-directed variant of the classic Ding-Lawrence algorithm~\citep{Ding2003}. Following Deigan\etal\cite{Deigan2009}, soft constraints are used to complement the free energy contributions of the classic Turner energy model with pseudo energy contributions resulting from reactivity profiles derived from SHAPE probing experiments.  Given a reactivity $R_i$ for a position $i$, we associate a free-energy bonus to unpaired positions, defined as
$$\Delta G(i) = m \log(R_i +1 )+b$$ 
using $m=1.3$ and  $b=-0.4$. Those values differ from those used by Deigan\etal\cite{Deigan2009}, and were adapted based on the rationale that lower absolute values for pseudo-energy bonuses increase the chance of structures being simultaneously supported by multiple conditions. Those pseudo energy contributions effectively direct our predictions towards a subset of structures that are in good agreement with probing data.

In order to infer recurrent candidates across different samples, sampled structure sets are agglomerated while keeping track of the reagent/condition of origin for each structure, and clustered using the Mini Batch k-Means algorithm~\citep{Sculley2010} (\CL{}) based on a number of clusters determined by an iterative heuristics described in further details below. 
It takes as input a dissimilarity matrix, whose entries consist of the base pair distance between two sampled structures.
\CL{} was chosen as it requires less computational resources than the classic k-means algorithm, yet performs equally as well as an extensive collection including  both agglomerative (affinity propagation) and hierarchical (Ward, Diana, McQuitty) clustering algorithms (data not shown).

The number of clusters is a critical parameter of the \CL{} algorithm. It should, at the same time, remain small enough to ensure reproducibility, while being sufficiently large to discriminate outliers and ensure consistency within each cluster. We iteratively increase the number of clusters, deeming it appropriate if increasing its value leads to:
\begin{enumerate} 
	\item Splitting a significant cluster into two highly similar clusters, or; 
	\item Creating a poorly populated (outlier) cluster, while keeping highly similar clusters for each significant cluster from the previous iteration.
\end{enumerate} 
In practice, clusters are seldom exactly preserved across successive iterations, so we consider two clusters to be \Def{highly similar} if their Maximum Expected Accuracy (MEA) centroid structures~\citep{Lu2009}, differ by less than $\delta$ base pairs.
The $\delta$ parameter defaults to $1$, allowing the identification of clusters across successive generation in the presence of a minor variation.
Moreover, define the (pseudo-)Boltzmann probability of a structure $S$, generated for an experimental condition $d$ as part of a sampled set $\mathcal{S}_d$, as
\[\mathbb{P}_d(S) = \frac{e^{-E_d(S)/kT}}{\mathcal{Z}^*_d} \text{, with } \mathcal{Z}^*_d := \sum_{S'\in \mathcal{S}_d} e^{-E_d(S')/kT}\]
where $E_d(S)$ is the pseudo free-energy assigned to the structure $S$ within the experimental condition $d$.
We consider a cluster to be \Def{significantly populated} if the cumulative probability of its structures exceeds a predefined threshold $\epsilon$. 
The value of $\epsilon$ was set to $\#{\sf Datasets}/3$, ensuring that up to three clusters are deemed significantly populated, and used as our primary candidates.

Based on these definitions, our iterative heuristic consists in running \CL over an increasing number $\NumClust$ of clusters, starting with $\NumClust=2$, until a \Def{stopping criterion} is satisfied, that is either: 1) Two significantly populated clusters have associated centroids which are highly similar; or 2) Centroid structures of significantly populated clusters from the previous iteration are highly similar to those of the current iteration.

>>>>

\maketitle
\begin{abstract}
The manual production of reliable RNA structure models from chemical probing experiments benefits from the integration of information derived from multiple protocols and reagents. However, the interpretation of multiple probing profiles remains a complex task, hindering the quality and reproducibility of modeling efforts.

@@

Reagents used within \Rev{selective 2 -hydroxyl acylation analyzed by primer extension (SHAPE)~\cite{Wilkinson2006} protocols} represent a popular class of small molecules. They react with the 2’-hydroxyl of flexible ribose~\cite{McGinnis2012}, although the exact properties observed by SHAPE remain the object of ongoing investigations~\cite{McGinnis2012,Sexton2017,Hurst2018,Mlynsky2018,Frezza2019,Busan2019}.
As ribose flexibility is proportional to the degree of freedom of the nucleotide, it is assumed that SHAPE reagents discriminate nucleotides in stable interactions from others. Different SHAPE reagents are endowed with different dynamics, that can help differentiate Watson-Crick base pairs from more dynamic tertiary contacts~\cite{Gherghe2008,Steen2012,Rice2014,Busan2019}. Other chemical probes, harbouring different chemical reactivities, have been developed. Amongst the most popular, DiMethyl Sulfate (DMS) is a small molecule that methylates N1 of adenines and N3 of cytosine if they are not involved in a hydrogen bond. Similarly, 1-cyclohexyl-3-(2-morpholinoethyl)carbodiimide metho-p-toluene sulfonate (CMCT) reacts on the Watson-Crick face of guanosine (N1) and uracyl (N3) when not engaged in a hydrogen bond~\cite{Ehresmann1987,Brunel2000}. These two reagents therefore mostly reveal Watson-Crick base pairing, but also any other type of contact involving the same edge. This diversity of probes not only increases coverage of the different positions and structural contexts in an RNA, but it also provides different qualitative information\Rev{, including in an \emph{in vivo} setting~\cite{Wang2019}}. It has been widely used since the very early days of RNA structure probing~\cite{Moazed1986, Ehresmann1987, Romaniuk1988, Butcher1994, Brunel2000,  Cordero2015, Somarowthu2015, Gross2017}.

@@

Computationally, the reactivity of a nucleotide is typically used as a proxy \Rev{to assess the unpaired nature of individual nucleotides}. The last couple of decades have nevertheless seen a series of paradigm shifts in the ways probing information is integrated, somehow mirroring the evolution of {\em ab initio} methods for secondary structure prediction. The seminal work of Mathews~\cite{Mathews2004} used cutoffs to transform reactivity values into hard constraints. Depending on the reagent/enzyme, significantly unreactive or reactive positions were forced to either paired or remain unpaired respectively within predicted structures. However, such \emph{hard} constraints turned out to be very sensitive to the choice of the cutoff, easily leading to artificially unpaired predictions or unsatisfiable sets of constraints. With the advent of SHAPE probing, subsequent methods lifted the requirement of thresholds and integrated reactivities directly as pseudo-energies, supplementing the Turner nearest-neighbor energy model~\cite{Turner2010} within an (pseudo) energy minimization scheme~\cite{Deigan2009,Zarringhalam2012,Washietl2012}. Such methods can be seen as optimizing a tradeoff between the thermodynamic stability of a model, assessed by the Turner energy model, and its compatibility with the reactivity profile. As a consequence, popular packages for secondary structure prediction, such as \Software{RNAStructure}~\cite{Mathews2004} or the \Software{Vienna} package~\cite{Lorenz2011}, now routinely accept SHAPE-like reactivity profiles as input of their main structure prediction methods. \Rev{Besides free-energy optimization, such methods notably} include the joint folding of aligned RNAs~\cite{Lavender2015}, partition function, statistical sampling and Maximal Expected Accuracy predictions~\cite{Spasic2017}\ldots

However, independent computational predictions for the same RNA using probing data obtained with different probes often yield substantially different models, none of which are fully consistent with all the probing data. 
It follows that, while theoretically informative, a multiprobing strategy often leaves a user with different models, from which one cannot objectively decide. In order to identify the best fit with all the data, researchers resort to very intuitive and manual methods such as projecting the results obtained with one probe onto the different models obtained using the constraints from another probe~\cite{Herbreteau2005,James2008,Weill2004}. At the end of the modeling process, the modeler is often left with several alternatives, all of which may appear equally consistent with the probing data. Thus, we have developed a new modeling procedure that jointly takes into account multiple probing data, and ultimately yields a small collection of secondary structure models.

@@

\subsection*{The \OurTool{} method}
We introduce the Integrative Probing Analysis of Nucleic Acids Empowered by Multiple Accessibility Profiles (\OurTool{}), a novel approach that integrates the signals produced by various chemical probes \Rev{to predict one or several secondary structure models for a given RNA.}

@@

	\caption{\OurTool{} workflow: \OurTool{} takes as input an RNA sequence with profiling data, denoted by reactivities, from various experimental conditions. \OurTool{} proceeds, first,  with a stochastic sampling that results into samples of predicted secondary structure. The data-driven predicted structures are then gathered in one sample, serving as input for the clustering step. \OurTool{} proceeds, then, with an iterative clustering that ends once \Rev{a stopping criterion is} reached. This step allows to identify the adequate number of clusters $k$ to be considered. The $k$ resulting clusters are then \Rev{represented} by their centroid structures. Clusters figuring on the 2D-Pareto frontier are considered to be optimal and subsequently their corresponding centroids are reported as the predicted structure through \OurTool.}\label{fig:approach}
\end{figure}

\paragraph{Sampling the pseudo-Boltzmann ensembles.} Our method, summarized in Figure~\ref{fig:approach}, takes as input a set $\mathcal{D}$ of probing experiments, each materialized by a reactivity profile. It starts by producing (multi)sets of representative structures for each of the reactivity profiles using a SHAPE-directed variant of the classic Ding-Lawrence algorithm~\citep{Ding2003}. Following Deigan\etal\cite{Deigan2009}, \Def{soft constraints} are used to complement the free energy contributions of the classic Turner energy model with pseudo energy contributions resulting from the reactivity derived from a probing experiment.  Given a reactivity $r_i$ for a position $i$ in a probing $d$, we associate a free-energy bonus to unpaired positions, defined as
$$\Delta G_d(i) = m \log(r_i +1 )+b$$ 
using $m=1.3$ and  $b=-0.4$. \Rev{Those values were halved in comparison to those recommended by Deigan\etal\cite{Deigan2009}, following a grid search optimization on the Cordero dataset, and based on the rationale that lower absolute values for pseudo-energy bonuses increase the expected overlap between pseudo-Boltzmann ensembles.} Those pseudo energy contributions effectively guide our predictions towards a subset of structures that are in good agreement with probing data. For each condition $d$ in $\mathcal{D}$, we use the soft constraints framework~\cite{Lorenz2016} of the \Software{RNAfold} software~\cite{Lorenz2011} to produce a random (multi)set $\mathcal{S}_d$ of $\NumStruct$ structures in the pseudo-Boltzmann ensemble.

\Rev{\paragraph{Clustering across conditions.} In order to infer recurrent conformations across sampled sets, \OurTool{} agglomerate structure (multi)sets while keeping track of their condition of origin, and \Def{clusters} with respect to the \Def{base-pair distance}, the number of base pairs differing between two structures. A clustering algorithm then partitions the (multi)set of sampled structures into \emph{clusters}, (multi)sets of structures such that the accumulated sum of distances over clusters is minimized.

Among the many available options, we chose the Mini Batch $\NumClust$-Means algorithm~\citep{Sculley2010} (MBkM), implemented in the \Software{scikit-learn} pacakage~\cite{Pedregosa2012}, which requires less computational resources than the classic $\NumClust$-means algorithm, yet performed similarly in preliminary studies as an extensive collection of both agglomerative (affinity propagation) and hierarchical (Ward, Diana, McQuitty) clustering algorithms. A dissimilarity matrix, presenting the pairwise base-pair distance between structures, is precomputed and fed to the clustering algorithm.

Any cluster $C$ output by the clustering is a multiset of structures, each labeled with its origin condition of $\mathcal{D}$.
The \Def{cluster probability of a structure feature} $f$ (base pair or unpaired base) within a cluster $C$ is then defined as 
$$ \mathbb{P}_C(f) = \frac{\sum_{\substack{S\in \mathcal{C}\\\text{s.t. } f\in S}} e^{-E(S)/RT}}{\sum_{\substack{S'\in \mathcal{C}}} e^{-E(S')/RT}} $$ 
where $R$ represents the Boltzmann constant, $E(S)$ is the Turner free-energy, and $\mathcal{C}$ is the (non-redundant) set of structures in $C$. From those probabilities we define the \Def{centroid structure} of a cluster as its Maximum Expected Accuracy (MEA) structure, computed efficiently following Lu\etal~\citep{Lu2009}.
